---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
devtools::install_github("tomaszaba/ipccheckr")
library(ipccheckr)
```

# `ipccheckr`:Toolkit for Performing IPC Acute Malnutrition-related Data Checks

<!-- badges: start -->
[![R-CMD-check](https://github.com/tomaszaba/ipccheckr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tomaszaba/ipccheckr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/tomaszaba/ipccheckr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/tomaszaba/ipccheckr?branch=main)
<!-- badges: end -->

## Background

IPC AMN is a global tool that uses a set o protocols to classify geographical areas into a 5-phases of severity based on the prevalence of acute malnutrition among children aged 6-59 months. Its main objective is to provide actionable information for decision-makers. It is not a data collection method _per se_, therefore, it relies on evidence generated through other methods (representative surveys, screening, sentinel sites, etc). Before any evidence is used an IPC AMN analysis, checks must performed to ascertain the quality of data, following the minimum quality standards set in the protocol, hence its reliability. As such, only evidence that pass the checks are allowed to be used in an IPC analysis exercise.

## Why `ipccheckr`?

The typical analysis workflow is overly complex, tiring and tedious, time-consuming and not not reproducible. Oftentimes, due to demand, the workflow needs to be implemented on a pile of data in absolute short time, making the process prone to human errors due to fatigue. In the workflow, data “checker” needs to travel from software to software as follows: SPSS software (mostly) to process data and export to Excel or CSV to then import it into Emergency Nutrition Assessments (ENA) for SMART software for to run the quality. From this stage downstream, the workflow needs to be done one-by-one for the number of survey area a data set may hold: run the quality check, then transfer the summary results into a summary table (Excel spreadsheet or other). Then, run the prevalence analysis (also one-by-one), and then transfer the results into a spreadsheet and so on and so forth. Sometimes, a fourth tool is used to correct for a likely overestimation of acute malnutrition prevalence based on MUAC when there are excess of younger children (6:59 months) over older children (24:59 months) in the sample. Therefore, the `ipccheckr` was developed simply to make the workflow simpler, joyful and reproducible thanks to its array of handy functions.

## What does `ipccheckr` do?

In a nutshell, the array of exported functions span from sample size checkers, data processors, quality checkers, prevalence calculators as well as some handy functions to render formatted and presentable output tables on the two latter groups of functions.

### Useful workflow with `ipccheckr` for data quality checks
```{r mermaid-diagram, echo=FALSE, fig.align='center'}
library(DiagrammeR)

mermaid(
diagram = "
graph TB;
  A[Start] --> B[Get acquainted with the data at your disposal];
  B --> C[Check if the sample size requirements were met];
  C --> D{Met?}
  D --> |NO| E(End);
  D --> |YES| F{Indicator?}
  F --> G{WFHZ};
  G --> H[Process age];
  H --> I[Process WFHZ data];
  I --> J[Check Plausibility];
  J --> K(End);
  F --> L{MFAZ};
  L --> M[Process age];
  M --> N[Process MUAC data];
  N --> O[Check Plausibility]
  O --> P(End);
  F --> Q{MUAC};
  Q --> R[Process MUAC data];
  R --> S[Check Plausibility];
  S --> T(End);
  
  style A fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style B fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style C fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style D fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style E fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style F fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style G fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px; 
  style H fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style I fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style J fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style K fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style L fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style M fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style N fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style O fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style P fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style Q fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style R fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style S fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px;
  style T fill:#FDFEFE, stroke:#715D58, stroke-width:1.5px
  
  
", width = 800, height = 1000 )
```

### On the prevalence analysis

After the data quality checks workflow, you will now be in position to decide whether or not you can proceed to the prevalence analysis. The workflow is quite simple: you use the data returned by the data processors in the above workflow. `ipccheckr` can compute acute malnutrition prevalence based one:

1. Weight-for-height z-scores - guide
2. Mid-Upper-Arm Circumference (using the absolute values) - guide
4. Combined prevalence - guide
3. MUAC-for-age z-score - guide

**IMPORTANT**: Please note that `ipccheckr` is still highly experimental and is undergoing a lot of development. Hence, any functionalities described below have a high likelihood of changing interface or approach as we aim for a stable working version.

## Installation

`ipccheckr` is not yet on CRAN but you can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tomaszaba/ipccheckr")
```

Then load to in memory with
```{r example}
library(ipccheckr)
```

# Citation

If you were enticed to using `ipccheckr` package and found it useful, please cite using the suggested citation provided by a call to `citation` function as follows:

```{r citation}
citation("ipccheckr")
```
