---
title: "Checking if IPC Acute Malnutrition minimum sample size requeriments were met"
author: "Tomás Zaba"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking if IPC Acute Malnutrition minimum sample size requeriments were met}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

library(ipccheckr)
```
Evidence on acute malnutrition prevalence used in IPC Acute Malnutrition (IPC AMN) scale can come from different sources and collected in different ways, be it representative surveys, screenings or even community-based surveillance system (known as sentinel sites). The IPC set minimum parameters on sample size for each of the above sources. Details on each of them can be read from the [IPC Manual version 3.1 ](https://www.ipcinfo.org/ipcinfo-website/resources/ipc-manual/en/). 

If you are using `ipcchecker` package utilities for the purpose of IPC, then you can easily check if you data meets the criteria defined in the manual by calling the `check_sample_size()` function.

To demonstrate its usage, two `ipccheckr` built-in sample data are used in this step-by-step guidance. Let's start with the sample data called `anthro.01`. You can inspect the data set as follows:

```{r usage.1, echo=TRUE, eval=FALSE}
head(anthro.01)
```

and this will return: 

```{r usage.1a, echo=FALSE, eval=TRUE}
head(anthro.01)
```

`anthro.01` contains data on district level SMART surveys from an anonymized location. You can check this data set's dictionary by calling `?anthro.01` in R's console. 

Now that you have inspected your data, you are now set to check if your data meets the IPC AMN criteria on the sample size perspective. To achieve that on the first data set you do the following: 

```{r usage.2, echo = TRUE, eval = FALSE}
check_sample_size(
  df = anthro.01,
  .group = cluster,
  .data_type = "survey"
)
```

Or you can also choose to chain the data object with the function using the pipe operator: 
```{r usage.2a, echo = TRUE, eval = FALSE}
anthro.01 |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

Either way, the returned output will be: 
```{r usage.2b, echo = FALSE, eval = TRUE}
anthro.01 |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

A table (of type `tibble`) is returned with three columns. Column **groups** counts the number of unique cluster ID's in the data set, so that if you survey has 40 cluster, then the group column will have that value. The **n_obs** columns stores the number of rows in the data set with any value. That is the number of children in your data set. Lastly, **meet_ipc** tells you whether or not the values in column one and two satisfies the IPC AMN sample size requirements for surveys. 

So far, and as you may have already realized, the above implementation only works for one survey data. However, you are most likely working with a multi-areas (provinces, counties, district, livelihoods, etc) survey data and, therefore, you want to check if the sample size requirements were met in each survey area. Using the above approach will summarise everything into one line, as shown above. You can achieve your desire by implementing as follows: 

```{r usage.2c, echo = TRUE, eval = FALSE}
## Load the `dplyr` package ----
library(dplyr)

## Use the `group_by() function ----
anthro.01 |> 
  group_by(area) |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

This will return the same type of output as before, but this time with multiple rows, depending on the number of unique districts (in the case of `anthro.01`). 

```{r usage.2d, echo = FALSE, eval = TRUE, message = FALSE}
## Load the `dplyr` library ----
library(dplyr)

## Use the `group_by()` function ----
anthro.01 |> 
  group_by(area) |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

If you are working on different data source, you will still implement the same way as shown before, however you would just change the input of `.data_type` to tell `check_sample_size()` on what basis it should evaluate if the conditions were met. As you saw, in previous demonstrations `.data_type` was set to `"survey"`. If your data come from a screening, then you will have to set `.data_type` to `"screening"`. If it comes from a sentinel site, you will have to set it to `"ssite"`. 

For this, another `ipccheckr` built-in sample data called `anthro.04` is used. This data was from a sentinel sites from an anonymized location.

You can inspect the data as follows:

```{r usage.3, echo=TRUE, eval=FALSE}
head(anthro.04)
```

and this will return: 

```{r usage.3a, echo=FALSE}
head(anthro.04)
```

Then you check if the IPC AMN requirements were met as follows: 
This time we want to display the first 20 rows: 

```{r usage.3b, echo=TRUE, eval=FALSE}
head(anthro.04, 20)
```

This will return:

```{r usage.3c, echo=FALSE}
head(anthro.04, 20)
```

Then you can do: 

```{r usage.3d, echo=TRUE, eval=FALSE}
## Load the `dplyr` library ----
library(dplyr)

## Use check_sample_size() function with .data_type set to "ssite" ----
anthro.04 |> 
  group_by(province) |> 
  check_sample_size(.group = cluster, .data_type = "ssite")

```

This will return: 

```{r usage.3e, echo=FALSE}
library(dplyr)

anthro.04 |> 
  group_by(province) |> 
  check_sample_size(.group = cluster, .data_type = "ssite")
```

## So what? 

So this is how you would check if your data did meet the IPC AMN requirements depending on the source or data type. Normally, this would be your first step in the `ipccheckr` workflow as it is in the IPC AMN data checks workflow. That means that if your data did meet the minimum requirements, you would then advance in the workflow to check the quality your data and then compute the prevalence, otherwise your workflow would end here. 

A step-by-step vignette for performing data quality checks as well computing prevalence is also available in this package. 
